---
title: "Sewanee's Birds"
author: 'Jillian Hall'
output: flexdashboard::flex_dashboard
---

```{r}
################ Loading in packages and Data #######################
library(readr)
library(DT)
library(ggplot2)
library(dplyr)
library(leaflet)
library(tinytable)
##################### DATA ##################
#loading in data but not using first row as names 
birds_data <- read_csv("birds_data.csv",
    col_names = FALSE)

#renaming the columns names 
birds_data<-birds_data%>%rename('building'=X4,'species'=X2,'identifier'=X3)

#making a vector for buildings
building<-c('mcclurg','quintard','tennessee ave','snowden','dupont library','fitwell','beihl','mccrady','memorial cross') 

#making a vectors of the latitude of the buildings 
lat<-c(35.20322,35.19755,35.20028,35.20332,35.20472,35.20286,35.20411,35.20284,35.19232)

#making a vector of the longitude of the buildings
lng<-c(-85.92029,-85.92522,-85.92205,-85.91758,-85.91785,-85.92076,-85.92134,-85.91811,-85.93706)

#adding all these vectors to gether to make a dataframe 
locations<-data.frame(building,lat,lng)
#grouping by building to find out how many birds were seen at each one 
total_observations<-birds_data%>%
  group_by(building)%>%
  summarise(total=n())

#grouping by building then finding the number of unique species are at each one 
number_species<-birds_data%>%
  group_by(building)%>%
  summarise(n_species=n_distinct(species))

#merging the two summaries into one data set by building 
totals<-merge(total_observations,number_species,by='building')

#then merging this new data set with the lat and lng we created earlier
totals<-merge(totals,locations,by='building')

#now making a templeate for r to paste into my map as a popup 
totals$popup_text<-paste0('Location: ', totals$building,'<br/>',
                          'Birds total: ',totals$total,'<br/>',
                          'Number ofspecies:',totals$n_species,'<br/>')
# finding the total for each species
speciestotal<-birds_data%>%group_by(species)%>%summarise(count=n())

```

##### **Bird observations were collected across nine locations on Sewanee’s campus to explore how bird communities vary across space. Birds were identified either visually or through vocalizations and recorded using a Shiny data entry app. This dashboard highlights patterns in species presence and detection across Sewanee's campus.**

## Column {.tabset data-width="800,"}

### **Map of Observation Locations**
Map shows the locations of observations along with site specific data. When markers are selected, data on the number of unique species and total number of birds seen is displayed. The color and size of the maker corresponds to the number of species present at that site.

```{r echo=FALSE, fig.height=7}

totals%>%leaflet()%>%
  addTiles()%>%
  setView(lng=-85.9211,lat=35.2031,zoom=14.5)%>%
  addCircleMarkers(
    lat = ~lat,
    lng = ~lng,
    radius = ~n_species,
    color = ~colorNumeric(c('lightskyblue','midnightblue'),n_species)(n_species),
    fillOpacity = 0.9,
    popup = ~popup_text
  )
```

### **Observation Counts By Species**

```{r fig.height=8, fig.width=7}
ggplot(speciestotal,aes(y=species,x=count))+geom_bar(stat='identity')+labs(title='Number of Observations of Each Species',caption='displays how many birds within each species across all locations were seen or heard.')+theme_minimal()

```

### **Identification Mode**

```{r fig.width=8}
#making data set with summaries of total number of times a bird was id-ed with audio or visuals
buildingid<-birds_data%>%group_by(identifier,building)%>%summarise(total=n(),.groups = "drop")

#making a grouped bar graph of the toatl number of observations made with each identification method
ggplot(buildingid,aes(x=building,y=total,fill=identifier))+geom_bar(position='dodge',stat = 'identity')+labs(caption='The number of birds identified at each site by either sight or sound.',x='Building')

```

## Column {data-width="150"}

### **Most Observed Species**

```{r fig.height=8, fig.width=4}
tt(speciestotal%>%arrange(desc(count))%>%slice(1:5))
```

### **Total Observations**

```{r fig.height=2}
tt(total_observations%>%summarise(Total_Overall=sum(total)))
```
